Dispersion Distance and the Matter Distribution of the Universe in Dispersion Space

### 2015-06-04 writeup/dm_space_distortions.tex
update to not-FRB caveat

However, redshift is not the only measure of radial distance.   Standard
candles and standard rulers can be used to estimate distance using brightness
and angular size respectively.  In the same way a \emph{standard ping}, a brief
broadband radio impulse, can be used to estimate radial distance because of the
characteristic $\lambda^2$ dispersion of electromagnetic signals along the
line-of-sight as a function of wavelength.  This technique, routinely done with pulsars in the Galaxy, can be used to define
a \emph{dispersion distance} to cosmological sources.  Intriguingly, the dispersion of fast radio bursts (FRB), a new class
of millisecond duration radio transients first reported by \citet{2007Sci...318..777L}, has been measured with high significance. 
These bursts are the archetype of a standard ping with 
10 reported
\citep{2012MNRAS.425L..71K,2013Sci...341...53T,
2014ApJ...790..101S, 2014ApJ...792...19B, 
2015MNRAS.447..246P,2015ApJ...799L...5R} to date and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
Dispersion is characterized by a line-of-site integral of the free electron density known the dispersion measure
(DM).
FRB events are observed to have dispersion measures of order
$\sim\!\!1000 {\rm pc} / {\rm cm}^3$, greatly in excess of
the galactic expectation of order $\sim\!\!100\,{\rm pc} / {\rm cm}^3$ (depending on
galactic latitude)
\cite{2014arXiv1412.4829D}.
They  thus appear to be of extra-galactic origin, with their
dispersion produced either by the
inter-galactic medium (IGM) \citep{2013Sci...341...53T} surrounding galaxies or the
environment of the source
\citep{2014ApJ...785L..26L,2015arXiv150101341P,2015arXiv150505535C}.
If the dispersion is dominated by electrons along the line-of-sight,
then the average DM--distance relation can be modelled
\citep{2014ApJ...797...71Z,2004MNRAS.348..999I, 2003ApJ...598L..79I}
and the sources appear to be at
cosmological distances of of order $\sim\!\!{\rm Gpc}$. Such cosmological distances permit
the study of clustering of FRB sources in three dimensions with an expectation that these events trace the
large-scale structure of the Universe on linear scales. Note that although we
frame our discussion in terms of FRBs, other bright radio transients that fluctuate on timescales less than the typical lag time of might also be used as standard pings.

>>>>

However, redshift is not the only measure of radial distance.   Standard
candles and standard rulers can be used to estimate distance using brightness
and angular size respectively.  In the same way a \emph{standard ping}, a brief
broadband radio impulse, can be used to estimate radial distance because of the
characteristic $\lambda^2$ dispersion of electromagnetic signals along the
line-of-sight as a function of wavelength.  This technique, routinely done with pulsars in the Galaxy, can be used to define
a \emph{dispersion distance} to cosmological sources.  Intriguingly, the dispersion of fast radio bursts (FRB), a new class
of millisecond duration radio transients first reported by \citet{2007Sci...318..777L}, has been measured with high significance. 
These bursts are the archetype of a standard ping with 
10 reported
\citep{2012MNRAS.425L..71K,2013Sci...341...53T,
2014ApJ...790..101S, 2014ApJ...792...19B, 
2015MNRAS.447..246P,2015ApJ...799L...5R} to date and total rate of bursts
been estimated to be $3.3^{+5}_{-2.5} \times 10^3$
per sky per day \cite{2015arXiv150500834R}. 
Dispersion is characterized by a line-of-site integral of the free electron density known the dispersion measure
(DM).
FRB events are observed to have dispersion measures of order
$\sim\!\!1000 {\rm pc} / {\rm cm}^3$, greatly in excess of
the galactic expectation of order $\sim\!\!100\,{\rm pc} / {\rm cm}^3$ (depending on
galactic latitude)
\cite{2014arXiv1412.4829D}.
They  thus appear to be of extra-galactic origin, with their
dispersion produced either by the
inter-galactic medium (IGM) \citep{2013Sci...341...53T} surrounding galaxies or the
environment of the source
\citep{2014ApJ...785L..26L,2015arXiv150101341P,2015arXiv150505535C}.
If the dispersion is dominated by electrons along the line-of-sight,
then the average DM--distance relation can be modelled
\citep{2014ApJ...797...71Z,2004MNRAS.348..999I, 2003ApJ...598L..79I}
and the sources appear to be at
cosmological distances of of order $\sim\!\!{\rm Gpc}$. Such cosmological distances permit
the study of clustering of FRB sources in three dimensions with an expectation that these events trace the
large-scale structure of the Universe on linear scales. We note that while our discussion here is framed in terms of FRBs, other bright radio transients that fluctuate on timescales less than the typical differential lag of $\sim 1s/100 \,{\rm MHz} $ might also be useful as standard pings.
